A meta-analysis of mental rotation ability in the first years of life

### 2022-05-05 code/analysis.Rmd
Update references
Incorporate gender difference analysis

abstract: |
  Mental rotation, the cognitive process of moving an object in mind to predict how it looks in a new orientation, is tightly coupled to intelligence, learning and educational achievement [@shepard1971; @hegarty1999; @johnson2005]. On average, males solve mental rotation tasks slightly faster than females [@linn1985; @voyer1995; @voyer2011; @maeda2013]. When such behavioral differences emerge during development, however, remains poorly understood [@lauer2019a]. Here we analyzed effect sizes derived from 59 experiments conducted in 1,798 infants aged 3 to 16 months. We robustly found that male infants looked longer at novel rotated objects than female infants. The effect size of this average difference was small and did not change with age. These findings indicate that gender differences in mental rotation are already present in the first months of life.

@@

```{r, setup, include=FALSE}
# Load packages
library(papaja)
library(here)
library(scales)
library(tidyverse)
library(magrittr)
library(metafor)
library(brms)
library(bayestestR)
library(tidybayes)
library(cowplot)

@@

# Read sheets
screening <- readxl::read_excel(excel_file, sheet = "screening")
included <- readxl::read_excel(excel_file, sheet = "included", na = "NA")

@@

    # Add unique experiment identifier
    experiment = str_c(year, article, group, sep = ", "),

@@

To confirm the gender difference between males and females we set up another Bayesian meta-analysis, this time focusing on the looking-time contrasts between all-male and all-female groups reported within each article by the original authors.
Our additional analysis revealed a meta-analytic effect size of g = 0.20, 95% Crl [0.03, 0.38] (Supplementary Fig. 1) and a probability of this effect being greater than zero of 98.9%.
The heterogeneity of effect sizes was \$sigma\$2Experiment = 0.01, 95% CrI [0.00, 0.07], at the experiment level, and \$sigma\$2Article = 0.02, 95% CrI [0.00, 0.10] at the article level.
This means that 56.1% of the total variance can be attributed to differences between articles and 43.1% to differences within articles.

@@

@@

>>>>

abstract: |
  Mental rotation, the cognitive process of moving an object in mind to predict how it looks in a new orientation, is tightly coupled to intelligence, learning and educational achievement [@shepard1971; @hegarty1999; @johnson2005]. On average, males solve mental rotation tasks slightly faster than females [@linn1985; @voyer1995; @voyer2011; @maeda2013]. When such behavioral differences emerge during development, however, remains poorly understood [@lauer2019a]. Here we analyzed effect sizes derived from 59 experiments conducted in 1,798 infants aged 3 to 16 months. We robustly found that male infants recognized novel rotated objects more reliably than female infants. The effect size of this average difference was small and did not change with age. These findings indicate that gender differences in mental rotation are already present in the first months of life.

@@

```{r, setup, include=FALSE}
# Load packages
library(papaja)
library(here)
library(readxl)
library(scales)
library(tidyverse)
library(magrittr)
library(metafor)
library(brms)
library(bayestestR)
library(tidybayes)
library(cowplot)

@@

# Read sheets
screening <- read_excel(excel_file, sheet = "screening")
included <- read_excel(excel_file, sheet = "included", na = "NA")
gender <- read_excel(excel_file, sheet = "gender_differences", na = "NA")

@@

    # Add unique experiment identifier
    experiment = str_c(year, article, group_long, sep = ", "),

@@

```{r, gender_differences}
# Extract relevant effect sizes
gender %>%
  # Remove redundant samples
  filter(!redundant) %>%
  # Add columns
  mutate(

    # Add unique experiment identifier
    experiment = str_c(year, article, group_long, sep = ", "),

    # Re-code non-significant f values as an effect of 0
    f = as.numeric(f),
    f_assumed = ifelse(is.nan(f), 0, f),

    # Add d from two-samples t test (Lakens, 2013)
    d_t = t * sqrt(1 / female_n + 1 / male_n),

    # Add d from ANOVA F value via conversion to a t value
    d_f = sqrt(f_assumed) * sqrt(1 / female_n + 1 / male_n),

    # Add d from mean difference and pooled standard deviation
    mean_diff = mean_diff_males_mean - mean_diff_females_mean,
    sd_diff_pooled_numerator = (male_n - 1) * (mean_diff_males_sd^2) +
      (female_n - 1) * (mean_diff_females_sd^2),
    df = male_n + female_n - 2,
    sd_diff_pooled = sqrt(sd_diff_pooled_numerator / df),
    d_diff = mean_diff / sd_diff_pooled,

    # Add d from one-sample t test of novelty preference scores
    mean_nov_pref = novelty_pref_males_mean - novelty_pref_females_mean,
    sd_nov_pref_pooled_numerator = (male_n - 1) * (novelty_pref_males_sd^2) +
      (female_n) * (novelty_pref_females_sd^2),
    sd_nov_pref_pooled = sqrt(sd_nov_pref_pooled_numerator / df),
    d_nov_pref = mean_nov_pref / sd_nov_pref_pooled,

    # Choose one type of outcome variable for each experiment
    di = case_when(
      # 1. If d was reported directed
      !is.na(d) ~ d,
      # 2. If a t test was reported
      !is.na(d_t) ~ d_t,
      # 3. If ANOVA was reported
      !is.na(d_f) ~ d_f,
      # 4. If the difference between means and their SDs were reported
      !is.na(d_diff) ~ d_diff,
      # 5. If novelty preference scores and their SDs were reported
      !is.na(d_nov_pref) ~ d_nov_pref
    ),
    # Keep track which type of outcome measure was chosen for each article
    di_type = case_when(
      !is.na(d) ~ "d",
      !is.na(d_t) ~ "d_t",
      !is.na(d_f) ~ "d_f",
      !is.na(d_diff) ~ "d_diff",
      !is.na(d_nov_pref) ~ "d_nov_pref"
    ) %>%
      factor(levels = c("d", "d_t", "d_f", "d_diff", "d_nov_pref")),

    # Apply small sample correction using Hedges' exact method
    # See http://dx.doi.org/10.20982/tqmp.14.4.p242
    j = exp(lgamma(df / 2) - log(sqrt(df / 2)) - lgamma((df - 1) / 2)),
    gi = di * j,

    # Compute standard error of Cohen's d, using harmonic mean of sample sizes
    ni = female_n + male_n,
    nhi = (female_n * male_n) / ni,
    vi = (df * 2 * (1 + (gi^2 * nhi * 0.5))) / ((df - 2) * nhi) - (gi^2 / j^2),
    sei = sqrt(vi)
  ) %>%
  # Remove experiments that didn't provide an effect size
  filter(!is.na(gi)) -> dat_gender

# Overview of the different effect sizes
dat_gender %>%
  select(
    article,
    group_long,
    gi,
    di,
    di_type,
    d,
    d_t,
    d_f,
    d_diff,
    sei
  ) %>%
  print(n = Inf)

# Specify priors
prior_c <- c(
  set_prior("normal(0, 1)", class = "b"),
  set_prior("cauchy(0, 0.3)", class = "sd")
)

# Run prior predictive simulation
res_prior_gender <- brm(
  gi | se(sei) ~ 0 + Intercept + (1 | article / experiment),
  data = dat_gender,
  prior = prior_c,
  sample_prior = "only",
  save_pars = save_pars(all = TRUE),
  chains = n_chains,
  iter = n_iter,
  warmup = n_warmup,
  cores = n_chains,
  control = list(adapt_delta = 0.99),
  file = here("results", "models", "res_prior_gender")
)
summary(res_prior_gender)
# plot(res_prior_gender)

# Run Bayesian multilevel model
res_brm_gender <- update(
  res_prior_gender,
  sample_prior = FALSE,
  file = here("results", "models", "res_brm_gender")
)
summary(res_brm_gender)
# plot(res_brm_gender)

# Get posterior probability mass > 0 for the meta-analytic effect
(hyp_brm_gender <- hypothesis(res_brm_gender, "Intercept > 0")$hypothesis)

# # Summarise fixed effects (incl. Bayes factors)
# describe_posterior(
#   res_brm_gender,
#   centrality = "mean", ci = 0.95, ci_method = "ETI",
#   test = c("p_direction", "rope", "bayesfactor"),
#   rope_range = c(-0.1, 0.1)
# )

# Extract posterior draws for the meta-analytic effects
spread_draws(res_brm_gender, `b_.*`, `sd_.*`, regex = TRUE, ndraws = NULL) %>%
  # Compute variances and ICC
  mutate(
    intercept = b_Intercept,
    sigma_article = sd_article__Intercept,
    sigma_experiment = `sd_article:experiment__Intercept`,
    sigma2_article = sigma_article^2,
    sigma2_experiment = sigma_experiment^2,
    sigma2_total = sigma2_article + sigma2_experiment,
    icc = sigma2_article / sigma2_total,
    .keep = "unused"
  ) -> draws_brm_gender

# Summarize as mean and 95% credible intervals
(summ_brm_gender <- as.list(mean_qi(draws_brm_gender)))
```

To confirm the gender difference between males and females we set up another Bayesian meta-analysis, this time focusing on the looking-time contrasts between all-male and all-female groups reported within each article by the original authors.
Our additional analysis revealed a meta-analytic effect size of $g = `r summ_brm_gender$intercept`$, 95% Crl [`r summ_brm_gender$intercept.lower`, `r summ_brm_gender$intercept.upper`] (Supplementary Fig. 1) and a probability of this effect being greater than zero of `r print_perc(hyp_brm_gender$Post.Prob)`.
The heterogeneity of effect sizes was $\sigma_{\text{Experiment}}^2 = `r summ_brm_gender$sigma2_experiment`$, 95% CrI [`r summ_brm_gender$sigma2_experiment.lower`, `r summ_brm_gender$sigma2_experiment.upper`], at the experiment level and $\sigma_{\text{Article}}^2 = `r summ_brm_gender$sigma2_article`$, 95% CrI [`r summ_brm_gender$sigma2_article.lower`, `r summ_brm_gender$sigma2_article.upper`] at the article level. Therefore, approximately `r print_perc(1 - summ_brm_gender$icc)` of the heterogeneity between effect sizes was attributable to differences between experiments and `r print_perc(summ_brm_gender$icc)` was attributable to differences between articles.

@@

Blah blah blah

# References

<div id="refs"></div>

\newpage

@@

```{r, figs1, include=TRUE, fig.height=10, fig.width=12, fig.cap="(ref:figs1-caption)"}
# Get posterior draws for the effect in each experiment
epred_draws_brm_gender <- epred_draws(res_brm_gender, dat_gender, ndraws = NULL) %>%
  mutate(experiment_f = factor(experiment))

# Create forest plot
dat_gender %>%
  # Make sure the plot will be ordered alphabetically by experiment IDs
  arrange(experiment) %>%
  mutate(
    experiment_f = fct_rev(fct_expand(factor(experiment), "model")),
    # Show article labels only for the first experiment per article
    article = if_else(article == lag(article, default = ""), "", article),
    # Compute frequentist confidence intervals
    ci_lb = gi - qnorm(0.05 / 2, lower.tail = FALSE) * sqrt(vi),
    ci_ub = gi + qnorm(0.05 / 2, lower.tail = FALSE) * sqrt(vi),
    ci_print = print_ci(gi, ci_lb, ci_ub)
  ) %>%
  # Prepare plotting canvas
  ggplot(aes(x = gi, y = experiment_f)) +
  geom_vline(xintercept = seq(-3, 3, 0.5), color = "grey90") +
  # Add article and group labels as text on the left
  geom_text(aes(x = -9.9, label = article), hjust = 0) +
  geom_text(aes(x = -7.1, label = group_long), hjust = 0) +
  # Add experiment-specific effect sizes and CIs as text on the right
  geom_text(aes(x = 3.7, label = print_num(gi)), hjust = 1) +
  geom_text(aes(x = 4.4, label = print_num(ci_lb)), hjust = 1) +
  geom_text(aes(x = 5.1, label = print_num(ci_ub)), hjust = 1) +
  # Add Bayesian highest posterior density intervals for each experiment
  stat_interval(
    aes(x = .epred),
    data = epred_draws_brm_gender,
    alpha = .8,
    point_interval = "mean_qi",
    .width = c(0.5, 0.95),
  ) +
  scale_color_grey(
    start = 0.85, end = 0.65,
    labels = as_mapper(~ scales::percent(as.numeric(.x)))
  ) +
  # Add experiment-specific effect sizes and CIs as dots with error bars
  geom_linerange(aes(xmin = ci_lb, xmax = ci_ub), size = 0.35) +
  geom_point(aes(size = ni), shape = 22, fill = "white") + # Or (1 / vi)?
  # Add posterior distribution for the meta-analytic effect
  stat_halfeye(
    aes(x = intercept, y = -1),
    draws_brm_gender,
    point_interval = "mean_qi",
    .width = c(0.95)
  ) +
  annotate(
    "text",
    x = c(-9.9, 3.7, 4.4, 5.1),
    y = -0.5,
    label = c(
      "Three-level model",
      print_num(summ_brm_gender$intercept),
      print_num(summ_brm_gender$intercept.lower),
      print_num(summ_brm_gender$intercept.upper)
    ),
    hjust = c(0, 1, 1, 1),
    fontface = "bold"
  ) +
  # Add headers for each of the for columns
  annotate( # Just to make the grid lines disappear
    "rect",
    xmin = -Inf,
    xmax = Inf,
    ymin = nrow(dat_gender) + 0.5,
    ymax = Inf,
    fill = "white"
  ) +
  annotate(
    "text",
    x = c(-9.9, -7.1, 0, 3.7, 4.4, 5.1),
    y = nrow(dat_gender) + 1.5,
    label = c(
      "Article",
      "Experiment",
      "Effect size plot",
      "Effect size",
      "Lower",
      "Upper"
    ),
    hjust = c(0, 0, 0.5, 1, 1, 1),
    fontface = "bold"
  ) +
  # Styling
  coord_cartesian(
    ylim = c(-1.5, nrow(dat_gender) + 2), expand = FALSE, clip = "off"
  ) +
  scale_x_continuous(breaks = seq(-3, 3, 0.5)) +
  annotate("segment", x = -3.3, xend = 3.3, y = -1.5, yend = -1.5) +
  labs(
    x = expression("Hedges'" ~ italic("g")),
    size = "Sample size",
    color = "CrI level"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(family = "Helvetica", color = "black"),
    axis.text.y = element_blank(),
    axis.ticks.x = element_line(colour = "black"),
    axis.title.x = element_text(hjust = 0.67, margin = margin(b = -15)),
    axis.title.y = element_blank(),
    panel.grid = element_blank(),
    text = element_text(family = "Helvetica")
  )

# Save the plot
ggsave(here(figures_dir, "forest_gender.pdf"), width = 12, height = 8)
```

(ref:figs1-caption) **Meta-analysis of gender differences.** A Bayesian three-level meta-analysis provided evidence for a small gender difference in mental rotation performance between male and female infants. White squares indicate the effect sizes (Hedges’ $g$) for gender difference in all individual experiments and black lines their 95% confidence intervals. Note that for experiments that observed a non-significant gender difference and did not specify the exact size of this effect, we assumed an effect size of $g = 0.00$. Gray bars indicate the 95% and 50% Bayesian credible interval (CrI) based on a Bayesian three-level random-effects model. Note that these intervals got shrunk towards the meta-analytic effect size because of partial pooling, that is, the three-level model regularizing the impact of experiments with small sample sizes and/or unrealistically large effect sizes. The last line shows the meta-analytic effect size from the three-level model (black dot) together with its 95% CrI (black line) and its posterior distribution (gray curve).
